---
title: "Projects"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 1
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
devtools::install_github("hadley/emo")
```  
 
A collection of personal projects that i have worked on. A data(tail) of everything from basic cleaning data, to simple data exploration to predictive analytics. Feel free to share your feedback or even contribute (Github) `r emo::ji("smile")`.


# NYC Schools Survey

2021 | 04 | 01 Last compiled: `r Sys.Date()`

[**This project**](https://rickyboshe.github.io/NYC-Schools/) analyzes the relationship between high school performance and the demographic makeup of schools in New York City. Additionally, a survey of parents, students and teachers perceptions on the quality of schools is analyzed and compared with actual performance. 

<center>
```{r, echo=FALSE, out.width="75%", fig.align="center"}
knitr::include_url("images/Graph.html", height = "350px")
```
</center>

<br><br>

# My second post 

2018 | 7 | 23 Last compiled: `r Sys.Date()`

**Coming Soon watch this space** 
