---
title: 'NYC Real Estate'
output:
  html_document:
    collapsed: false
    number_sections: false
    toc_depth: 3
    code_folding: hide
---

<br></br>
<center>
#  Predicting Real Estate Prices in New York City `r emo::ji("house_with_garden")`
</center>


New York city is known to be one of the most expensive cities in the world when it comes to real estate. The city has 5 different **boroughs**, with each having its unique socioeconomic profile that helps determine the real estate prices. 

<iframe title="New York City Boroughs" aria-label="Map" id="datawrapper-chart-S0nLu" src="https://datawrapper.dwcdn.net/S0nLu/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="669"></iframe>

<script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(e){if(void 0!==e.data["datawrapper-height"]){var t=document.querySelectorAll("iframe");for(var a in e.data["datawrapper-height"])for(var r=0;r<t.length;r++){if(t[r].contentWindow===e.source)t[r].style.height=e.data["datawrapper-height"][a]+"px"}}}))}();
</script>


Using data shared by the [New York City department of Finance](https://www1.nyc.gov/site/finance/taxes/property-rolling-sales-data.page), this project looks to analyze housing costs for the last 12 months and use regression models to predict prices based on indicators.
 
```{r setup, include=FALSE}
#Loading packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)
library(ggcorrplot)
library(scales)
library(lubridate)
library(jsonlite)
library(readxl)
library(DataExplorer)
library(broom)
library(kableExtra)
library(jtools)
library(pander)

library(webshot2)

#Load data
manhattan<-read_excel("C:/Users/ricky/Documents/Datasets/NYC Property Sales/rollingsales_manhattan.xls", skip = 4)
brooklyn<-read_excel("C:/Users/ricky/Documents/Datasets/NYC Property Sales/rollingsales_brooklyn.xls", skip = 4)
staten<-read_excel("C:/Users/ricky/Documents/Datasets/NYC Property Sales/rollingsales_statenisland.xls", skip = 4)
bronx<-read_excel("C:/Users/ricky/Documents/Datasets/NYC Property Sales/rollingsales_bronx.xls", skip = 4)
queens<-read_excel("C:/Users/ricky/Documents/Datasets/NYC Property Sales/rollingsales_queens.xls", skip = 4)

```



## Data Manipulation
The data is found in 4 distinct excel files. Read them to the environment and merge them in a single dataframe, making it easier to handle. 

### Data Cleaning
```{r cleaning, include=TRUE}
#The boroughs are coded (manhattan=1, bronx=2, brooklyn=3, queens=4, staten=5)
nyc<-bind_rows(manhattan,bronx,brooklyn,queens,staten)

#Recode the boroughs
nyc<-nyc%>%
  mutate(
    BOROUGH=case_when(
      BOROUGH=="1" ~ "Manhattan",
      BOROUGH=="2" ~ "Bronx",
      BOROUGH=="3" ~ "Brooklyn",
      BOROUGH=="4" ~ "Queens",
      BOROUGH=="5" ~ "Staten Island"
      )
    )

#remove the individual dataframes
rm(bronx, brooklyn, manhattan, queens, staten)

#Normalize the column names (lower column names and remove space)
colnames(nyc)<-str_to_lower(colnames(nyc))%>%
  str_replace_all("\\s", "_")%>%
  str_to_title(colnames(nyc))

#some of the properties were exchanged between family members (i.e. filter with a threshold of 10,000$)
summary(nyc$Sale_price)
nyc<-nyc%>%
  filter(Sale_price>10000)

#Some properties have square footage of 0, which is unlikely in reality. Remove
summary(nyc$Gross_square_feet)
nyc<-nyc%>%
  filter(Gross_square_feet>0)

#Initial removal of variables that are not of interest
nyc<-nyc%>%
  dplyr::select(-5,-6,-7,-8,-10)
```

The initial data cleaning and manipulation helps set the data up ready for exploration and followed by analysis i.e. removing missing observations and exploring the data. 


### Data Exploration
```{r explore, include=TRUE, fig.align='center', fig.width=8, message=FALSE}
plot_intro(nyc, ggtheme = theme_bw())

sum(is.na(nyc$Year_built))
nyc<-nyc%>%
  drop_na(Year_built)

table(nyc$Borough)

#Plot continuous and discrete values
plot_histogram(nyc, ggtheme = theme_bw(), ncol = 2)
plot_bar(nyc, maxcat = 5, 
         ggtheme = theme_bw())

#Keep only family dwellings only
nyc<-nyc%>%
  filter(Building_class_category=="01 ONE FAMILY DWELLINGS" | Building_class_category=="02 TWO FAMILY DWELLINGS" | Building_class_category=="03 THREE FAMILY DWELLINGS")%>%
  mutate(Building_class_category=
           case_when(
             Building_class_category=="01 ONE FAMILY DWELLINGS" ~ "One Family Dwellings",
             Building_class_category=="02 TWO FAMILY DWELLINGS" ~ "Two Family Dwellings",
             Building_class_category=="03 THREE FAMILY DWELLINGS" ~ "Three Family Dwellings"
             )
         )

table(nyc$Borough)

plot_bar(nyc, maxcat = 5,
         ncol = 2,
         ggtheme = theme_bw())

```
Very few missing observations (14), and they seem to be from the *Year Built* column. We can go ahead and remove them since they make up just 0.04% of the data.


You can also observe that most variables have very huge outliers that skew their charts positively. Going forward, we might need to deal with the outliers. One potential way is by keeping just residential dwellings (building codes can be found [here](https://www1.nyc.gov/assets/finance/jump/hlpbldgcode.html)). Manhattan will see a considerable drop in observations (<span style="color: red;">76%</span>) while the least drop in observations was for Staten Island (<span style="color: red;">2%</span>).  


### Data Outliers
```{r outliers, include=TRUE}
#Residential Units
quartiles <- quantile(nyc$Residential_units)
# 75% minus 25% = interquartile range 
iqr <- quartiles[[4]] - quartiles[[2]]
# Outlier boundaries
lower_bound <- quartiles[[2]] - (1.5 * iqr)
upper_bound <- quartiles[[4]] + (1.5 * iqr)

# Isolate outlier(s)
res.outliers <- nyc%>% 
  filter(Residential_units > upper_bound | Residential_units< lower_bound)

#Land
quartiles <- quantile(nyc$Land_square_feet)
iqr <- quartiles[[4]] - quartiles[[2]]
lower_bound <- quartiles[[2]] - (1.5 * iqr)
upper_bound <- quartiles[[4]] + (1.5 * iqr)

land.outliers <- nyc%>% 
  filter(Land_square_feet > upper_bound | Land_square_feet< lower_bound)

#Price
quartiles <- quantile(nyc$Sale_price)
iqr <- quartiles[[4]] - quartiles[[2]]
lower_bound <- quartiles[[2]] - (1.5 * iqr)
upper_bound <- quartiles[[4]] + (1.5 * iqr)

price.outliers <- nyc%>% 
  filter(Borough=="Bronx")%>%
  filter(Sale_price > upper_bound | Sale_price< lower_bound)

#Bronx has two duplicate outliers, remove them
nyc<-nyc%>%
  filter(Sale_price!=87400000)

#remove duplicates
nyc<-nyc%>%
  distinct()
```


## Modeling
To successfully model real estate prices we must select the best predictor variables (independent variables) that will best explain the prices. 


### Factor selection
As OLS assumes no multicollinearity between independent variables, a correlation matrix helps us identify variables that are strongly correlated and thus removable. In our case, *Total Units* is highly correlated with *Residential units* and mildly correlated with *Gross Square Feet*. We therefore remove *Total Units* from the equation. 
```{r factor, include=TRUE, fig.align='center', fig.width=8}
#Recode columns to proper data types
nyc<-nyc%>%
  mutate(Borough=as.factor(Borough),
         Neighborhood=as.factor(Neighborhood),
         Tax_class_at_time_of_sale=as.factor(Tax_class_at_time_of_sale),
         Zip_code=as.factor(Zip_code),
         Building_class_at_time_of_sale=as.factor(Building_class_at_time_of_sale),
         Building_class_category=as.factor(Building_class_category))
nyc$Address<-NULL
nyc$Neighborhood<-NULL
nyc$Zip_code<-NULL
nyc$Tax_class_at_present<-NULL

#Check for multicollinearity
numnyc <- names(which(sapply(nyc, is.numeric)))
corr <- cor(nyc[,numnyc], use = 'pairwise.complete.obs')
p3<-ggcorrplot(corr, lab = TRUE)
p3

#Total units has strong relationship with Residential units, so i shall remove it
nyc$Total_units<-NULL

#Visualize relationship between sale price and gross size
nyc%>%ggplot(aes(x=Gross_square_feet, y=Sale_price, color=Borough))+
  geom_point()+
  theme_bw()+
  geom_smooth(method = "lm", se = FALSE)+
  theme(legend.position = "none")+
  facet_wrap(~Borough, ncol = 2, scales = "free")
```



It is important to model each borough individually as they are different in terms of socio-economic profile. It is logical that each borough would have its own price drivers. We can do this using *nesting* techniques.

### Regression
Taking an overall regression for the full dataset, we see the model helps explain 43% of the observations significantly. As mentioned earlier, it would be wise to seperate observations by Borough and run seperate models for each Borough. 
```{r regression, include=TRUE, warning=FALSE, message=FALSE}
#Regression with two most important factors, Borough and gross size of property
nyc_fit<-lm(Sale_price~Borough+Gross_square_feet, data=nyc)
summ(nyc_fit)

#Best model by Borough (Generate multiple linear models)
##First nest data by the category
nyc_nest<-nyc%>%
  group_by(Borough)%>%
  nest()

##Second run a liner regression across all categories, using a mapping function
nyc_nest<-nyc_nest%>%
  mutate(linear_model=map(.x= data, 
                          .f= ~lm(Sale_price~+Gross_square_feet, data=.)
                          ))

##Third select the broom function suitable
#a) Tidy the data
nyc_nest<-nyc_nest%>%
  mutate(tidy_coef=map(.x=linear_model,
                       .f=tidy, conf.int=TRUE))

#Fourth we Unnest
p_borough<-nyc_nest%>%
  dplyr::select(Borough, tidy_coef)%>%
  unnest(cols = tidy_coef)%>%
  filter(term=="Gross_square_feet")%>%
  arrange(estimate)

panderOptions('table.split.table', 300)
pander(p_borough, round=3, caption="<b>Change in Real Estate prices per increase in square foot by Borough</b>")

#a) glance the data (this has r-squared)
nyc_nest<-nyc_nest%>%
  mutate(glance_coef=map(.x=linear_model,
                       .f=glance))

#we Unnest
r_borough<-nyc_nest%>%
  dplyr::select(Borough, glance_coef)%>%
  unnest(cols = glance_coef)%>%
  dplyr::select(r.squared, adj.r.squared, p.value, AIC, BIC)%>%
  arrange(r.squared)

panderOptions('table.split.table', 300)
pander(r_borough, round=3, caption="<b>How well Real Estate prices can be explained by the model for each Borough</b>")

```


## Conclusion
Not surprising that Manhattan has the highest increase in real estate price for each increase square foot of the property. Bronx has the smallest increase. Manhattan also has the best model fit, meaning 56% of the observation prices can be explained by the model. While for Brooklyn it is just 19% of the observations. This means there might be other price drives that we did not catch in our dataset. 